Estill 2016 reanalysis

### 2023-12-18 figs/ART_estill2016.Rmd
fix anno

# Annotation
anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group",
  "Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

## Probe sets

For each gene, extract out the probes.

```{r,probesets}

promoters <- grep("Prom",anno$Regulatory_Feature_Group)
promoters <- anno[promoters,]
gp <- anno[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
summary(unlist(lapply(sets,length)))

@@

TODO- add explanation

@@

ARRAY_SAMPLESHEET="GSE79257/GSE79257_Illumina_samplesheet.csv"
# only download it if it is not present on the system
if ( !file.exists(ARRAY_SAMPLESHEET ) ) {
    DLFILE=paste(ARRAY_SAMPLESHEET,".gz",sep="")
    download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE79nnn/GSE79257/suppl/GSE79257_Illumina_samplesheet.csv.gz",
        destfile = DLFILE)
    gunzip(DLFILE, overwrite = TRUE )
}

@@

# exclude sex chromosomes
keep <- !(featureNames(mSetSw) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
mSetFlt <- mSetSw[keep,]
head(mSetFlt)
dim(mSetFlt)

@@

head(top_nat_vs_fh$dma)
 head(top_nat_vs_fh$dmr)
top_nat_vs_fh$comp
top_nat_vs_fh$cgi
write.table(top_nat_vs_fh$dma,file="estill_nat_vs_fh.tsv",sep="\t",quote=FALSE)

@@

head(top_nat_vs_fz$dma)
head(top_nat_vs_fz$dmr)
top_nat_vs_fz$comp
write.table(top_nat_vs_fz$dma,file="estill_nat_vs_fz.tsv",sep="\t",quote=FALSE)

@@

@@

head(top_fh_vs_fz$dma)
head(top_fh_vs_fz$dmr)
top_fh_vs_fz$comp
write.table(top_fh_vs_fz$dma,file="estill_fh_vs_fz.tsv",sep="\t",quote=FALSE)

@@

head(top_nat_vs_fx$dma)
head(top_nat_vs_fx$dmr)
top_nat_vs_fx$comp
write.table(top_nat_vs_fx$dma,file="estill_nat_vs_fx.tsv",sep="\t",quote=FALSE)

@@

head(top_iui_vs_fx$dma)
head(top_iui_vs_fx$dmr)
top_iui_vs_fx$comp
write.table(top_iui_vs_fx$dma,file="estill_iui_vs_fx.tsv",sep="\t",quote=FALSE)

## Venn diagrams of the differential methylated probes for each contrast

First, we look at the similarity of DMPs altered by fresh and frozen procedures. The overlap is very large.
This indicates a high degree of similarity between the profiles.

```{r,venn1,fig.cap="Comparison of probes altered by fresh and frozen IVF procedures", fig.width = 8 ,fig.height = 8}

v1 <- list("FH up" = top_nat_vs_fh$dm_up , 
           "FZ up" = top_nat_vs_fz$dm_up ,
           "FH dn" = top_nat_vs_fh$dm_dn ,
           "FZ dn" = top_nat_vs_fz$dm_dn )
plot(euler(v1, shape = "ellipse"), quantities = TRUE)

Now to take a look at whether IUI is similar to IVF fresh and frozen. The overlap is just a large as 
FH vs FZ.

```{r,venn2,fig.cap="Comparison of probes altered by IUI, fresh and frozen IVF procedures", fig.width = 8 ,fig.height = 8}

v2 <- list("FH up" = top_nat_vs_fh$dm_up , 
           "FH dn" = top_nat_vs_fh$dm_dn ,
           "IUI up" = top_nat_vs_iui$dm_up ,
           "IUI dn" = top_nat_vs_iui$dm_dn )
plot(euler(v2, shape = "ellipse"), quantities = TRUE)

v3 <- list("FZ up" = top_nat_vs_fz$dm_up , 
           "FZ dn" = top_nat_vs_fz$dm_dn ,
           "IUI up" = top_nat_vs_iui$dm_up ,
           "IUI dn" = top_nat_vs_iui$dm_dn )
plot(euler(v3, shape = "ellipse"), quantities = FALSE)

## Spearman correlations of each contrast

While the above Venn diagrams are suggestive of similarity between contrasts, the best 
way to assess this is with correlation analysis. I have chosen Spearman as based on
the directional p-value ranking metric.

```{r,spearman, fig.width = 8 ,fig.height = 8}

mycontrasts <- list("top_nat_vs_fh"=top_nat_vs_fh, "top_nat_vs_fz"=top_nat_vs_fz, "top_nat_vs_fx"=top_nat_vs_fx ,
  "top_nat_vs_iui"=top_nat_vs_iui, "top_fh_vs_fz"=top_fh_vs_fz, "top_iui_vs_fx"=top_iui_vs_fx)

myrnks <- lapply(X = mycontrasts, FUN = myranks)
df <- join_all(myrnks,by="rn")
rownames(df) <- df$rn
df$rn=NULL
colnames(df) <- names(mycontrasts)
head(df)

mycors <- cor(df,method = "spearman")

my_palette <- colorRampPalette(c("darkred","red", "orange", "yellow","white"))(n = 25)
heatmap.2(mycors,scale="none",margin=c(10, 10),cexRow=0.8,trace="none",cexCol=0.8,
    col=my_palette,main="Spearman correlations")

mycors

>>>>

# Annotation
annofull <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno <- data.frame(annofull[,c("UCSC_RefGene_Name","Regulatory_Feature_Group",
  "Islands_Name","Relation_to_Island")])

@@

@@

ARRAY_SAMPLESHEET="GSE79257/GSE79257_Illumina_samplesheet.csv"
# only download it if it is not present on the system
if ( !file.exists(ARRAY_SAMPLESHEET ) ) {
    DLFILE=paste(ARRAY_SAMPLESHEET,".gz",sep="")
    options(timeout=1000000)
    download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE79nnn/GSE79257/suppl/GSE79257_Illumina_samplesheet.csv.gz",
        destfile = DLFILE)
    gunzip(DLFILE, overwrite = TRUE )
}

@@

# exclude sex chromosomes
keep <- !(featureNames(mSetSw) %in% annofull$Name[annofull$chr %in% c("chrX","chrY")])
mSetFlt <- mSetSw[keep,]
head(mSetFlt)
dim(mSetFlt)

@@

head(top_nat_vs_fh$dma)
 head(top_nat_vs_fh$dmr)
top_nat_vs_fh$comp
top_nat_vs_fh$cgi

saveRDS(top_nat_vs_fh,file="estill_nat_vs_fh.rds")

@@

head(top_nat_vs_fz$dma)
head(top_nat_vs_fz$dmr)
top_nat_vs_fz$comp

saveRDS(top_nat_vs_fz,file="estill_nat_vs_fz.rds")

@@

saveRDS(top_nat_vs_iui,file="estill_nat_vs_iui.rds")

@@

head(top_fh_vs_fz$dma)
head(top_fh_vs_fz$dmr)
top_fh_vs_fz$comp

saveRDS(top_fh_vs_fz,file="estill_fh_vs_fz.rds")

@@

head(top_nat_vs_fx$dma)
head(top_nat_vs_fx$dmr)
top_nat_vs_fx$comp

saveRDS(top_nat_vs_fx,file="estill_nat_vs_fx.rds")

@@

head(top_iui_vs_fx$dma)
head(top_iui_vs_fx$dmr)
top_iui_vs_fx$comp

saveRDS(top_iui_vs_fx,file="estill_iui_vs_fx.rds")
